Draft Final Report: Measles risk assessment, modelling, and benefit--cost analyses for New Zealand

### 2014-11-07 draftfinalreport.tex
Final edits but not compiling correctly

\begin{figure}
\begin{center}
\includegraphics{draftfinalreport-002}
\end{center}
\caption{Age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
\label{fig:ageinyears}
\end{figure}

@@

\begin{table}[hbtp]
\footnotesize
%latex.default(ttyr, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllr}
\hline\hline
\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Cases}\tabularnewline
\hline
1-5&0-2&None&$3$\tabularnewline
1-5&3-5&None&$1$\tabularnewline
1-5&6-17&None&$3$\tabularnewline
6-10&6-17&None&$4$\tabularnewline
1-5&18-24&None&$2$\tabularnewline
6-10&18-24&None&$1$\tabularnewline
1-5&25+&None&$5$\tabularnewline
6-10&25+&None&$3$\tabularnewline
\hline
\end{tabular}\end{center}\caption{Absolute number of measles cases in specific age and socio-economic deprivation categories from 2007-2014 with no ethnicity recorded}
\label{table:none}
\end{table}

The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}. Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses. However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see Table~\ref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table~\ref{table:percap}).

\begin{figure}
\begin{center}
\includegraphics{draftfinalreport-010}
\end{center}
\caption{Per capita rates of measles infections for specific age and New Zealand deprivation indices (NZDep) for 2007--2014, broken down by ethnicity (Table~\ref{table:percap} for details)}
\label{fig:percap}
\end{figure}

\begin{table}[hbtp]
\footnotesize
\caption{Numbers of measles cases, population sizes and per capita rates per 10,000 of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
%latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllrrr}
\hline\hline
\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Population}&\multicolumn{1}{c}{Cases 1}&\multicolumn{1}{c}{Per capita}\tabularnewline
\hline
1-5&0-2&Asian&$   6094$&$ 11$&$18.0505$\tabularnewline
6-10&0-2&Asian&$   6806$&$  8$&$11.7543$\tabularnewline
1-5&3-5&Asian&$   6094$&$  1$&$ 1.6410$\tabularnewline
6-10&3-5&Asian&$   6806$&$  0$&$ 0.0000$\tabularnewline
1-5&6-17&Asian&$  33918$&$ 11$&$ 3.2431$\tabularnewline
6-10&6-17&Asian&$  28905$&$  5$&$ 1.7298$\tabularnewline
1-5&18-24&Asian&$  22917$&$  3$&$ 1.3091$\tabularnewline
6-10&18-24&Asian&$  34107$&$  5$&$ 1.4660$\tabularnewline
1-5&25+&Asian&$  96357$&$ 10$&$ 1.0378$\tabularnewline
6-10&25+&Asian&$  98715$&$ 13$&$ 1.3169$\tabularnewline
1-5&0-2&European&$  57872$&$ 83$&$14.3420$\tabularnewline
6-10&0-2&European&$  45445$&$ 64$&$14.0830$\tabularnewline
1-5&3-5&European&$  57872$&$ 42$&$ 7.2574$\tabularnewline
6-10&3-5&European&$  45445$&$ 17$&$ 3.7408$\tabularnewline
1-5&6-17&European&$ 264330$&$219$&$ 8.2851$\tabularnewline
6-10&6-17&European&$ 182937$&$ 80$&$ 4.3731$\tabularnewline
1-5&18-24&European&$ 107649$&$ 34$&$ 3.1584$\tabularnewline
6-10&18-24&European&$ 117840$&$ 36$&$ 3.0550$\tabularnewline
1-5&25+&European&$1001916$&$ 78$&$ 0.7785$\tabularnewline
6-10&25+&European&$ 724317$&$ 51$&$ 0.7041$\tabularnewline
1-5&0-2&Maori&$  10003$&$ 18$&$17.9946$\tabularnewline
6-10&0-2&Maori&$  30104$&$ 48$&$15.9447$\tabularnewline
1-5&3-5&Maori&$  10003$&$  7$&$ 6.9979$\tabularnewline
6-10&3-5&Maori&$  30104$&$ 11$&$ 3.6540$\tabularnewline
1-5&6-17&Maori&$  40461$&$ 19$&$ 4.6959$\tabularnewline
6-10&6-17&Maori&$ 116640$&$ 92$&$ 7.8875$\tabularnewline
1-5&18-24&Maori&$  15360$&$  5$&$ 3.2552$\tabularnewline
6-10&18-24&Maori&$  48495$&$  8$&$ 1.6497$\tabularnewline
1-5&25+&Maori&$  71217$&$  2$&$ 0.2808$\tabularnewline
6-10&25+&Maori&$ 192729$&$  6$&$ 0.3113$\tabularnewline
1-5&0-2&MLA&$    728$&$  0$&$ 0.0000$\tabularnewline
6-10&0-2&MLA&$   1290$&$  3$&$23.2558$\tabularnewline
1-5&3-5&MLA&$    728$&$  0$&$ 0.0000$\tabularnewline
6-10&3-5&MLA&$   1290$&$  1$&$ 7.7519$\tabularnewline
1-5&6-17&MLA&$   2991$&$  0$&$ 0.0000$\tabularnewline
6-10&6-17&MLA&$   4539$&$  1$&$ 2.2031$\tabularnewline
1-5&18-24&MLA&$   1710$&$  0$&$ 0.0000$\tabularnewline
6-10&18-24&MLA&$   3078$&$  2$&$ 6.4977$\tabularnewline
1-5&25+&MLA&$   8028$&$  0$&$ 0.0000$\tabularnewline
6-10&25+&MLA&$  10335$&$  6$&$ 5.8055$\tabularnewline
1-5&0-2&Pacific&$   2093$&$  5$&$23.8892$\tabularnewline
6-10&0-2&Pacific&$  13124$&$ 58$&$44.1938$\tabularnewline
1-5&3-5&Pacific&$   2093$&$  0$&$ 0.0000$\tabularnewline
6-10&3-5&Pacific&$  13124$&$  3$&$ 2.2859$\tabularnewline
1-5&6-17&Pacific&$   8541$&$  5$&$ 5.8541$\tabularnewline
6-10&6-17&Pacific&$  51183$&$ 22$&$ 4.2983$\tabularnewline
1-5&18-24&Pacific&$   3972$&$  1$&$ 2.5176$\tabularnewline
6-10&18-24&Pacific&$  22098$&$  8$&$ 3.6202$\tabularnewline
1-5&25+&Pacific&$  18492$&$  2$&$ 1.0815$\tabularnewline
6-10&25+&Pacific&$  91533$&$ 11$&$ 1.2018$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:percap}
\end{table}

@@

The distribution of the meases cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
\begin{figure}
\begin{center}
\includegraphics{draftfinalreport-014}
\end{center}
\caption{Distribution of measles cases per category used in the final regression model}
\label{fig:histcasecat}
\end{figure}

@@

The significance of the different predictor variables can be seen in the ANOVA results (Table~\ref{table:anova}).
\par
\vspace{5mm} %5mm vertical space
\begin{table}
% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Fri Nov 07 22:00:40 2014
\begin{tabular}{lrrrrrr}
  \hline
 & Df & Deviance & Resid. Df & Resid. Dev & F & Pr($>$F) \\ 
  \hline
NULL &  &  & 39 & 1453.62 &  &  \\ 
  Age & 4 & 1304.20 & 35 & 149.43 & 183.58 & 0.0000 \\ 
  Ethnicity & 3 & 20.00 & 32 & 129.43 & 3.75 & 0.0285 \\ 
  NZDep & 1 & 10.70 & 31 & 118.74 & 6.02 & 0.0239 \\ 
  Age:Ethnicity & 12 & 81.91 & 19 & 36.83 & 3.84 & 0.0045 \\ 
   \hline
\end{tabular}\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using Equation~\ref{eq:reg} and a quasipoisson error structure}
\label{table:anova}
\end{table}
\vspace{5mm} %5mm vertical space
A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in Table~\ref{table:regression}.

\vspace{5mm} %5mm vertical space
% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Fri Nov 07 22:00:40 2014
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & -6.4674 & 0.1148 & -56.32 & 0.0000 \\ 
  Age3-5 & -0.9129 & 0.2054 & -4.44 & 0.0003 \\ 
  Age6-17 & -0.7618 & 0.1343 & -5.67 & 0.0000 \\ 
  Age18-24 & -1.5050 & 0.1937 & -7.77 & 0.0000 \\ 
  Age25+ & -2.9507 & 0.1608 & -18.35 & 0.0000 \\ 
  EthnicityAsian & 0.0530 & 0.3250 & 0.16 & 0.8721 \\ 
  EthnicityMaori & 0.2124 & 0.1994 & 1.07 & 0.2999 \\ 
  EthnicityPacific & 1.1603 & 0.2043 & 5.68 & 0.0000 \\ 
  NZDep6-10 & -0.2119 & 0.0855 & -2.48 & 0.0227 \\ 
  Age3-5:EthnicityAsian & -2.0315 & 1.3827 & -1.47 & 0.1581 \\ 
  Age6-17:EthnicityAsian & -1.0074 & 0.4717 & -2.14 & 0.0459 \\ 
  Age18-24:EthnicityAsian & -0.8311 & 0.5941 & -1.40 & 0.1779 \\ 
  Age25+:EthnicityAsian & 0.4211 & 0.4433 & 0.95 & 0.3541 \\ 
  Age3-5:EthnicityMaori & -0.3864 & 0.4096 & -0.94 & 0.3573 \\ 
  Age6-17:EthnicityMaori & -0.0855 & 0.2469 & -0.35 & 0.7328 \\ 
  Age18-24:EthnicityMaori & -0.5828 & 0.4484 & -1.30 & 0.2092 \\ 
  Age25+:EthnicityMaori & -1.0482 & 0.5242 & -2.00 & 0.0600 \\ 
  Age3-5:EthnicityPacific & -2.1316 & 0.8139 & -2.62 & 0.0169 \\ 
  Age6-17:EthnicityPacific & -1.4541 & 0.3347 & -4.35 & 0.0003 \\ 
  Age18-24:EthnicityPacific & -0.9827 & 0.5129 & -1.92 & 0.0705 \\ 
  Age25+:EthnicityPacific & -0.6127 & 0.4367 & -1.40 & 0.1767 \\ 
   \hline
\end{tabular}\caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
\label{table:regression}
\end{table}
\vspace{5mm} %5mm vertical space

\par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case. Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (Figure~\ref{fig:ageinyears}).

The majority of measles cases are among people of European ancestry (Table~\ref{table:percap} and Figure~\ref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001). Those people in NZDep levels 6-10 are under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes. In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected. The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}). This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:percap}). Interestingly the regression results suggest risk of measles in the 6--17 year age category was greater for Europeans and Maori.

@@

\subsection{Population immunity results}

@@

%%%%%%%%%%%%%%%%%%%

@@

Globally, analyses of the most complete and recent year of data available, 2012, suggest measles incidence highest  and vaccination coverage is lowest in less developed nations (Tables~\ref{table:incidence12} and ~\ref{table:cover12}, and Figures~\ref{fig:incidence12} and ~\ref{fig:cover12}). Immigration by non-New Zealanders (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:immigration12}. However, travel by New Zealanders is dominated by that to Australia \ref{table:nztravel12} and \ref{fig:nztravel12}. Together, these mean that the greatest travel location for New Zealanders and immigration origin is from Australia (\ref{table:immigration12} and \ref{fig:immigration12}. Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that those countries with greatest international travel, Australia and New Zealand, pose the greatest risk to New Zealand for measles importation. Though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand. These data are shown in Table~\ref{table:risk12}. The data for all the variables for each nation state and territories for 2012 are plotted in Figures~\ref{fig:immigration12}, ~\ref{fig:incidence12}, and ~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}. The breakdown by New Zealander and non-New Zealander travellers is in Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12} and Figures~\ref{fig:nzrisk12} and ~\ref{fig:imrisk12}.

@@

\begin{figure}
\begin{center}
\includegraphics{draftfinalreport-039}
\end{center}
\caption{Total international travel, 2012 }
\label{fig:travel12}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics{draftfinalreport-041}
\end{center}
\caption{Non-New Zealander international travel and immigration, 2012 }
\label{fig:immigration12}
\end{figure}

@@

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{nzers.pdf}
     \caption{Trends in international travel, determined by immigration arrival data and New Zealander travel departure destinations}
     \label{fig:travel}
\end{figure}

@@

The regression analyses suggest that age is a particularly strong risk factor for measles. This comes as no surpise to epidemiologists or health care providers. However, our analyses also highlight other groups that are at greater risk of measles. In particular, Pacific people are at greater risk, as are the less deprived (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age. Interpretation of these results must still be viewed with some caution, however, as there is very likely a spatial effect that might not be accounted for in these analyses. Additional data we have been provided by the Ministry of Health but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the NIR. A key issue with incorporating the spatial immunisation data has been the denominator data and the NIR data. The data suggest that census data and NIR data are recording children living in different DHBs to that in which they are vaccinated, or subsequent to the census or NIR data moving. This leads to some area census units having more than 100\% children vaccinated, with some many times more, and some with very low levels reportedly vaccinated. Another issue was how to deal with people of greater age than those recorded in the NIR. Thus, these analyses are possible future directions for this work and a focus of future data collection could be to better understand the link between denominator data and NIR data.

@@

Further analyses are required to see if the seasonal patterns in travel match seasonality in measles incidence in the countries from which most individuals come from when travelling to, or back to, New Zealand. However, the strong seasonality in travel around Christmas suggests this may be a time where extra effort and vigilence is required for measles vaccination and control efforts. The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (Table~\ref{table:immigration12} and Figure~\ref{fig:immigration12}). However, recently Australia was declared free of endemic measles and so the 2012 data must reflect imported measles there. However, other areas of high risk include the United Kingdom, where measles persists, and South, Southeast and East Asia, where measles is endemic (Table~\ref{table:risk12} and Figure~\ref{fig:risk12}).

>>>>

\begin{figure}
\begin{center}
\includegraphics{draftfinalreport-002}
\end{center}
\caption{Numbers and age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
\label{fig:ageinyears}
\end{figure}

@@

\begin{table}[hbtp]
\footnotesize
%latex.default(ttyr, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllr}
\hline\hline
\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Cases}\tabularnewline
\hline
1-5&0-2&None&$3$\tabularnewline
1-5&3-5&None&$1$\tabularnewline
1-5&6-17&None&$3$\tabularnewline
6-10&6-17&None&$4$\tabularnewline
1-5&18-24&None&$2$\tabularnewline
6-10&18-24&None&$1$\tabularnewline
1-5&25+&None&$5$\tabularnewline
6-10&25+&None&$3$\tabularnewline
\hline
\end{tabular}\end{center}\caption{Absolute number of measles cases with no ethnicity recorded for specific age and socio-economic deprivation (NZDep) categories from 2007-2014.}
\label{table:none}
\end{table}

The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}. Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses. However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see Table~\ref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table~\ref{table:percap}).

\begin{figure}
\begin{center}
\includegraphics{draftfinalreport-010}
\end{center}
\caption{Per capita rates of measles infections for specific age categories (0--2, 3--5, 6--17, 18--24, and 24+ years) and New Zealand deprivation indices (NZDep, 1--5 and 6--10) for 2007--2014, broken down by ethnicity (Table~\ref{table:percap} for details)}
\label{fig:percap}
\end{figure}

\begin{table}[hbtp]
\footnotesize
\caption{Numbers of measles cases (Cases), population sizes (Ppulation) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014}
%latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllrrr}
\hline\hline
\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Population}&\multicolumn{1}{c}{Cases 1}&\multicolumn{1}{c}{Per capita}\tabularnewline
\hline
1-5&0-2&Asian&$   6094$&$ 11$&$18.0505$\tabularnewline
6-10&0-2&Asian&$   6806$&$  8$&$11.7543$\tabularnewline
1-5&3-5&Asian&$   6094$&$  1$&$ 1.6410$\tabularnewline
6-10&3-5&Asian&$   6806$&$  0$&$ 0.0000$\tabularnewline
1-5&6-17&Asian&$  33918$&$ 11$&$ 3.2431$\tabularnewline
6-10&6-17&Asian&$  28905$&$  5$&$ 1.7298$\tabularnewline
1-5&18-24&Asian&$  22917$&$  3$&$ 1.3091$\tabularnewline
6-10&18-24&Asian&$  34107$&$  5$&$ 1.4660$\tabularnewline
1-5&25+&Asian&$  96357$&$ 10$&$ 1.0378$\tabularnewline
6-10&25+&Asian&$  98715$&$ 13$&$ 1.3169$\tabularnewline
1-5&0-2&European&$  57872$&$ 83$&$14.3420$\tabularnewline
6-10&0-2&European&$  45445$&$ 64$&$14.0830$\tabularnewline
1-5&3-5&European&$  57872$&$ 42$&$ 7.2574$\tabularnewline
6-10&3-5&European&$  45445$&$ 17$&$ 3.7408$\tabularnewline
1-5&6-17&European&$ 264330$&$219$&$ 8.2851$\tabularnewline
6-10&6-17&European&$ 182937$&$ 80$&$ 4.3731$\tabularnewline
1-5&18-24&European&$ 107649$&$ 34$&$ 3.1584$\tabularnewline
6-10&18-24&European&$ 117840$&$ 36$&$ 3.0550$\tabularnewline
1-5&25+&European&$1001916$&$ 78$&$ 0.7785$\tabularnewline
6-10&25+&European&$ 724317$&$ 51$&$ 0.7041$\tabularnewline
1-5&0-2&Maori&$  10003$&$ 18$&$17.9946$\tabularnewline
6-10&0-2&Maori&$  30104$&$ 48$&$15.9447$\tabularnewline
1-5&3-5&Maori&$  10003$&$  7$&$ 6.9979$\tabularnewline
6-10&3-5&Maori&$  30104$&$ 11$&$ 3.6540$\tabularnewline
1-5&6-17&Maori&$  40461$&$ 19$&$ 4.6959$\tabularnewline
6-10&6-17&Maori&$ 116640$&$ 92$&$ 7.8875$\tabularnewline
1-5&18-24&Maori&$  15360$&$  5$&$ 3.2552$\tabularnewline
6-10&18-24&Maori&$  48495$&$  8$&$ 1.6497$\tabularnewline
1-5&25+&Maori&$  71217$&$  2$&$ 0.2808$\tabularnewline
6-10&25+&Maori&$ 192729$&$  6$&$ 0.3113$\tabularnewline
1-5&0-2&MLA&$    728$&$  0$&$ 0.0000$\tabularnewline
6-10&0-2&MLA&$   1290$&$  3$&$23.2558$\tabularnewline
1-5&3-5&MLA&$    728$&$  0$&$ 0.0000$\tabularnewline
6-10&3-5&MLA&$   1290$&$  1$&$ 7.7519$\tabularnewline
1-5&6-17&MLA&$   2991$&$  0$&$ 0.0000$\tabularnewline
6-10&6-17&MLA&$   4539$&$  1$&$ 2.2031$\tabularnewline
1-5&18-24&MLA&$   1710$&$  0$&$ 0.0000$\tabularnewline
6-10&18-24&MLA&$   3078$&$  2$&$ 6.4977$\tabularnewline
1-5&25+&MLA&$   8028$&$  0$&$ 0.0000$\tabularnewline
6-10&25+&MLA&$  10335$&$  6$&$ 5.8055$\tabularnewline
1-5&0-2&Pacific&$   2093$&$  5$&$23.8892$\tabularnewline
6-10&0-2&Pacific&$  13124$&$ 58$&$44.1938$\tabularnewline
1-5&3-5&Pacific&$   2093$&$  0$&$ 0.0000$\tabularnewline
6-10&3-5&Pacific&$  13124$&$  3$&$ 2.2859$\tabularnewline
1-5&6-17&Pacific&$   8541$&$  5$&$ 5.8541$\tabularnewline
6-10&6-17&Pacific&$  51183$&$ 22$&$ 4.2983$\tabularnewline
1-5&18-24&Pacific&$   3972$&$  1$&$ 2.5176$\tabularnewline
6-10&18-24&Pacific&$  22098$&$  8$&$ 3.6202$\tabularnewline
1-5&25+&Pacific&$  18492$&$  2$&$ 1.0815$\tabularnewline
6-10&25+&Pacific&$  91533$&$ 11$&$ 1.2018$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:percap}
\end{table}

@@

The distribution of the meases cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
\begin{figure}
\begin{center}
\includegraphics{draftfinalreport-014}
\end{center}
\caption{Distribution of measles cases per category (Table~\ref{table:percap}) used in the final regression model (Equation~\ref{eq:reg})}
\label{fig:histcasecat}
\end{figure}

@@

The significance of the different predictor variables can be seen in the ANOVA results (Table~\ref{table:anova}).
\par
\vspace{5mm} %5mm vertical space
\begin{table}
% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Sat Nov 08 00:20:12 2014
\begin{tabular}{lrrrrrr}
  \hline
 & Df & Deviance & Resid. Df & Resid. Dev & F & Pr($>$F) \\ 
  \hline
NULL &  &  & 39 & 1453.62 &  &  \\ 
  Age & 4 & 1304.20 & 35 & 149.43 & 183.58 & 0.0000 \\ 
  Ethnicity & 3 & 20.00 & 32 & 129.43 & 3.75 & 0.0285 \\ 
  NZDep & 1 & 10.70 & 31 & 118.74 & 6.02 & 0.0239 \\ 
  Age:Ethnicity & 12 & 81.91 & 19 & 36.83 & 3.84 & 0.0045 \\ 
   \hline
\end{tabular}\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using Equation~\ref{eq:reg} and a quasipoisson error structure}
\label{table:anova}
\end{table}
\vspace{5mm} %5mm vertical space

A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in Table~\ref{table:regression}.

\vspace{5mm} %5mm vertical space
\begin{table}
% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Sat Nov 08 00:20:12 2014
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & -6.4674 & 0.1148 & -56.32 & 0.0000 \\ 
  Age3-5 & -0.9129 & 0.2054 & -4.44 & 0.0003 \\ 
  Age6-17 & -0.7618 & 0.1343 & -5.67 & 0.0000 \\ 
  Age18-24 & -1.5050 & 0.1937 & -7.77 & 0.0000 \\ 
  Age25+ & -2.9507 & 0.1608 & -18.35 & 0.0000 \\ 
  EthnicityAsian & 0.0530 & 0.3250 & 0.16 & 0.8721 \\ 
  EthnicityMaori & 0.2124 & 0.1994 & 1.07 & 0.2999 \\ 
  EthnicityPacific & 1.1603 & 0.2043 & 5.68 & 0.0000 \\ 
  NZDep6-10 & -0.2119 & 0.0855 & -2.48 & 0.0227 \\ 
  Age3-5:EthnicityAsian & -2.0315 & 1.3827 & -1.47 & 0.1581 \\ 
  Age6-17:EthnicityAsian & -1.0074 & 0.4717 & -2.14 & 0.0459 \\ 
  Age18-24:EthnicityAsian & -0.8311 & 0.5941 & -1.40 & 0.1779 \\ 
  Age25+:EthnicityAsian & 0.4211 & 0.4433 & 0.95 & 0.3541 \\ 
  Age3-5:EthnicityMaori & -0.3864 & 0.4096 & -0.94 & 0.3573 \\ 
  Age6-17:EthnicityMaori & -0.0855 & 0.2469 & -0.35 & 0.7328 \\ 
  Age18-24:EthnicityMaori & -0.5828 & 0.4484 & -1.30 & 0.2092 \\ 
  Age25+:EthnicityMaori & -1.0482 & 0.5242 & -2.00 & 0.0600 \\ 
  Age3-5:EthnicityPacific & -2.1316 & 0.8139 & -2.62 & 0.0169 \\ 
  Age6-17:EthnicityPacific & -1.4541 & 0.3347 & -4.35 & 0.0003 \\ 
  Age18-24:EthnicityPacific & -0.9827 & 0.5129 & -1.92 & 0.0705 \\ 
  Age25+:EthnicityPacific & -0.6127 & 0.4367 & -1.40 & 0.1767 \\ 
   \hline
\end{tabular}\caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
\label{table:regression}
\end{table}
\vspace{5mm} %5mm vertical space

\par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case. Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (see also Figure~\ref{fig:ageinyears}).

The majority of measles cases are among people of European ancestry (Table~\ref{table:percap} and Figure~\ref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001). Those people in NZDep levels 6-10 are under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes. In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected. The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}, Table~\ref{table:sero}, and Section~\ref{sub:popim}). This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:percap}). Interestingly the regression results suggest risk of measles in the 6--17 year age category was greater for Europeans and Maori.

@@

\subsection{Population immunity results}
\label{sub:popim}

@@

@@

Globally, analyses of the most complete and recent year of data available, 2012, suggest measles incidence highest  and vaccination coverage is lowest in less developed nations (Tables~\ref{table:incidence12} and ~\ref{table:cover12}, and Figures~\ref{fig:incidence12} and ~\ref{fig:cover12}). Immigration by non-New Zealanders (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:immigration12}. However, travel by New Zealanders is dominated by that to Australia (Tables~\ref{table:nztravel12} and \ref{fig:nztravel12}). Together, these mean that the greatest travel location for New Zealanders and immigration origin is from Australia (Table~\ref{table:immigration12} and Figure~\ref{fig:immigration12}. Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that those countries with greatest international travel, Australia and New Zealand, pose the greatest risk to New Zealand for measles importation. Though immigration from and travel to some Asian countries is lower, some Asian countries also poses a high risk of measles importation to New Zealand. These data are shown in Table~\ref{table:risk12}. The breakdown of risk of measles importation by New Zealander and non-New Zealander travellers is in Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12} and Figures~\ref{fig:nzrisk12} and ~\ref{fig:imrisk12}.

@@

\begin{figure}
\begin{center}
\includegraphics{draftfinalreport-039}
\end{center}
\caption{Non-New Zealander international travel and immigration, 2012 }
\label{fig:immigration12}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics{draftfinalreport-041}
\end{center}
\caption{Total international travel, 2012 }
\label{fig:travel12}
\end{figure}

@@

\begin{figure}[H]
     \begin{center}
     \includegraphics[width=0.9\textwidth]{nzers.pdf}
     \caption{Trends in international travel, determined by immigration arrival data and New Zealander travel departure destinations}
     \label{fig:travel}
\end{figure}

@@

The regression analyses suggest that age is a particularly strong risk factor for measles. This comes as no surpise to epidemiologists or health care providers. However, our analyses also highlight other groups that are at greater risk of measles. In particular, Pacific people are at greater risk per capita, as are the less deprived (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age. Interpretation of these results must still be viewed with some caution, however, as there is very likely a spatial effect that might not be accounted for in these analyses. Additional data we have been provided by the Ministry of Health but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the NIR. A key issue with incorporating the spatial immunisation data has been the denominator data and the NIR data. The data suggest that census data and NIR data are recording children living in different DHBs to that in which they are vaccinated, or subsequent to the census or NIR data moving. This leads to some area census units having more than 100\% children vaccinated, with some many times more, and some with very low levels reportedly vaccinated. Another issue was how to deal with people of greater age than those recorded in the NIR. Thus, these analyses are possible future directions for this work and a focus of future data collection could be to better understand the link between denominator data and NIR data.

@@

Further analyses are required to see if the seasonal patterns in travel match seasonality in measles incidence in the countries from which most individuals come from when travelling to, or back to, New Zealand. However, the strong seasonality in travel around the summer vacation in December suggests this may be a time where extra effort and vigilence is required for measles vaccination and control efforts. The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (Table~\ref{table:immigration12} and Figure~\ref{fig:immigration12}). However, recently Australia was declared free of endemic measles and so the 2012 data must reflect imported measles there. However, other areas of high risk include the United Kingdom, where measles persists, and South, Southeast and East Asia, where measles is endemic (Table~\ref{table:risk12} and Figure~\ref{fig:risk12}).

\subsection{Risk analysis summary}
\begin{itemize}
\item Risk of measles infection decreases significantly with age.
\item European ethnicity comprise the largest number of measles cases.
\item Pacific people are statistically more at risk per capita of measles infection.
\item There is some statistical support for those living in better socio-economic situations being at greater risk of measles.
\item There is some statistical support for Pacific and Asian children in the 6--17 year age categories being at lower risk per capita than European or Maori children.
\item The majority of cases are unvaccinated.
\item The majority of vaccine failures occur in those people which recieved single vaccinations around 1 year old.
\item Distribution of numbers of na\"{\i}ve among New Zealand is uneven, with the majority predictably in DHBs with larger urban areas.
\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
\item There may be seasonal changes in risk of measles importation, with travel numbers peaking in December.
\end{itemize}
